---
title: "About Variational Bayes for PPCA data"
author: "Eric PARENT & Pierre Gloaguen"
date: "the 30th of Mai 2022 , UMR MIA-Paris, AgroParisTech, INRA, Université Paris-Saclay"
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_document:
    code_folding: hide
    df_print: paged
    self_contained: yes
    number_sections: yes
    toc: yes
    theme: journal
    toc_float: yes
    toc_depth: 2
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE, cache = FALSE}

# knitr::opts_chunk$set(
# 	echo = TRUE,
# 	message = FALSE,
# 	warning = FALSE,
# 	cache = FALSE
# )
knitr::opts_chunk$set(
	eval = FALSE,
	message = FALSE,
	warning = FALSE,
	cache = FALSE,
	include = TRUE
)
```

# R codes and librairies {-}

All `R` codes can be unfolded, for instance, unfold the following chunk to see which libairies are used in this document:

```{r librairies}
rm(list=ls())
library(tidyverse) # For data manipulation and visualization
library(rjags)  # For MCMC sampling
library(ggmcmc) # To format jags output
library(extraDistr) # For the non standart t distribution
```


# A not-so-simple shrinkage PPCA model

Bhattacharya and Dunson consider the following linear normal model with a gamma process to shrink the variances of the columns of the latent matrix $\lambda$:
\begin{align*}
Y_i &
\overset{ind.}{\sim} \mathcal{N}_p(\Lambda\eta_i, \Sigma^{-1}), &
1\leq i \leq n 
\\
\Sigma &= 
\text{diag}(\sigma_j^{-2}), &
1\leq j \leq p 
\\
\sigma_j^{-2}|(a_{\sigma},b_{\sigma}) &
\sim \Gamma(a_{\sigma},b_{\sigma}) &
1\leq j \leq p
\\
\Lambda_{jh} &
\sim \mathcal{N}_{q}(0,\,\phi_{jh}^{-1}\tau_{h}^{-1}), &
1\leq j \leq p, 1 \leq h \leq {q}
\\
\phi_{jh} &
\sim \Gamma(\frac{\nu}{2},\frac{\nu}{2}), &
1\leq j \leq p, 1 \leq h \leq {q}\\
\tau_{h} &
= \prod_{l=1}^h \delta_l &
1\leq h \leq q
\\ 
\delta_1 &
\sim \Gamma(a_1,1) &
\\
\delta_h &
\sim \Gamma(a_2,1) &
\forall l \geq 2
\end{align*}

Recall that -- up to constant terms-- the logdensity of such a Normal multivariate distribution is:
$$\log[Y|\Lambda,\Sigma] = -\frac{n}{2}\left(\log\vert C\vert +\text{Tr}(C^{-1}\hat{S}_Y)\right)$$
with

$$ C= \Lambda\Lambda'+\Sigma\\
\hat{S}_Y=\frac{1}{n}\sum_{i=1}^n Y'_iY_i$$

In the case $q<<p$, because $\Sigma$ is diagonal, it is easier to compute the inverse of the variance-covariance matrix by $$C^{-1}=\Sigma^{-1}-\Sigma^{-1}\Lambda(I+ \Lambda' \Sigma^{-1}\Lambda)^{-1}\Lambda'\Sigma^{-1}$$


Some example data are generated in the following R code chunk.

```{r generation}
source("utils_PPCA_generating_data.R") # For true values
Y <- read.table("data_PPCA.txt", sep = ";") %>%
  as.matrix()
```


# Variational approach

## Variational family

We denote $Z$ the unknown with components $(\tau,\phi,\eta,\Lambda,\Sigma)$ and choose an independent family $q(Z) = \left(\prod_u q_{z_u}(Z_u) \right)$ to approximate $[Z|Y]$ such that :
\begin{align*}
\delta_h &
\sim q_{\delta_h}(\cdot)= \mathcal{G}(A^{\delta}_h,B^{\delta}_h), &
1\leq h \leq q 
\\
\phi_{jh} &
\sim q_{\phi_{jh}}(\cdot)= \mathcal{G}(A^{\phi}_{jh},B^{\phi}_{jh}),  &
1\leq j \leq p, 1\leq h \leq q
\\
\sigma^{-2}_{j} &
\sim q_{\sigma^{-2}_{j}}(\cdot) = \mathcal{G}(A^{\sigma}_j,B^{\sigma}_j), &
1\leq j \leq p 
\\
\eta_i &
\sim q_{\eta_i}(\cdot)= \mathcal{N}(M_{\eta_i}, V^{\eta_i}), & 
1\leq i \leq n
\\
\Lambda_j &
\sim q_{\Lambda_j }(\cdot)= \mathcal{N}(M_{\lambda_j},V^{\Lambda_j}), &
1\leq j \leq p, 
\end{align*}
where $\eta_i$ denotes the $i$-th row of yhe $n\times q$ matrix $\eta$, and $\Lambda_j$ denotes the $j$-th row of the $p\times q$ matrix $\Lambda$. In the following, all vectors are treated as column vectors.

## Evidence lower bound

To update parameters, we rely on the minimisation of the Kullback Liebler divergence
$KL= \mathbb{E}_q\left[\log\frac{q(Z)}{[Z|Y]}\right]$.
Equivalently we want to maximize the ELBO

$$\text{ELBO}= \mathbb{E}_q\left[\log[Y,\tau,\phi,\eta,\Lambda,\sigma^{-2}]\right]-\mathbb{E}_q\left[\log q(\tau,\phi,\eta,\Lambda,\sigma^{-2})\right]$$

Due to conditionnal independence the criterion can decomposed as 

\begin{align*}
\text{ELBO}=&\mathbb{E}_q\left[\log\left([Y|\eta,\Lambda,\Sigma]\right)\right]\\
& + \mathbb{E}_q\left[\log\left([\Sigma]\right)\right]\\
& + \mathbb{E}_q\left[\log\left([\eta]\right) \right]\\
& + \mathbb{E}_q\left[\log\left([\Lambda|\delta,\phi]\right)\right]\\
& + \mathbb{E}_q\left[\log\left([\phi]\right)\right]\\
& + \mathbb{E}_q\left[\log\left([\delta]\right)\right]\\
& -\mathbb{E}_q\left[\log q(\Lambda,\Sigma,\eta,\phi,\delta)\right]
\end{align*}

## Terms to integrate

Up to constant terms:
\begin{align*}
\log[Y|\Lambda,\Sigma,\eta]  &
= \frac{n}{2}\sum_{j = 1}^p \log \sigma^{-2}_j - \frac{1}{2}\sum_{j = 1}^p \sigma^{-2}_j\left(Y^{j} - \eta \Lambda_j\right)^T\left(Y^{j} - \eta \Lambda_j\right)
\\
&= -\frac{n}{2}\left(\log\vert \Sigma^{-1}\vert +\text{Tr}(\Sigma^{-1}\hat{S}_{Y,\eta,\Lambda})\right),
\end{align*}
with

$$ \hat{S}_{Y,\eta,\Lambda}=\frac{1}{n}\sum_{i=1}^n (Y_i-\Lambda\eta_i')(Y'_i-\eta_i\Lambda')\\
= \frac{1}{n} (Y'-\eta\Lambda')(Y-\lambda\eta').$$


Continuing (still up ton constant terms), we have for $\eta$:
$$\log[\eta] = -\frac{1}{2}\sum_{i = 1}^n \eta_i^T\eta_i = -\frac{1}{2}\text{Tr}\left(\eta'\eta\right).$$
Since $\Sigma$ is diagonal:

$$\log[\Sigma^{-1}] = \sum_{j=1}^p\left((a_\sigma -1)\log(\sigma_j^{-2}) -b_\sigma\sigma_j^{-2}\right)$$

Due to normal assumptions for $\Lambda_j$

$$\log\left([\Lambda_{j}|\tau_{h},\phi_{j,h}]\right)= \frac{1}{2}\sum_{h = 1}^q \left(\log \phi_{j,h} + \sum_{\ell = 1}^h\log \delta_\ell\right) - \frac{1}{2}\text{Tr}\left(\text{diag}\left(\phi_{j,h}\prod_{\ell = 1}^h\delta_\ell\right)\Lambda_{j}\Lambda_{j}^T \right).$$
Remains the with the prior for $\phi_{j,h}\;(1\leq j \leq p, 1\leq h \leq q)$:

$$\log[\phi_{jh}] = \left(\frac{\nu}{ 2} - 1\right)\log(\phi_{j,h}) -\frac{\nu}{ 2}\phi_{jh},$$
and $\delta_h\; (1\leq h \leq q)$

$$
\log[\delta_{h}] = \left\lbrace 
\begin{array}{lr}
(a_1 -1)\log(\delta_{1}) -b_{1} \delta_{1} & \text{ if } h = 1,\\
(a_2 -1)\log(\delta_{2}) -b_{2} \delta_{2} & \text{ else.}
\end{array}
\right.
$$


<!-- Wrapping up all the previous log density terms -->


<!-- \begin{align*} -->
<!-- \text{ELBO}=& -\mathbb{E}_q\left[\log q(\Lambda,\Sigma,\eta,\phi,\delta)\right]\\ -->
<!-- & +\mathbb{E}_q\left[\frac{n}{2}\left(\sum_{j=1}^p \log\sigma_j^{-2} \right)-0.5\sum_{i=1}^n tr\left(\Sigma^{-1} (Y_i-\Lambda\eta_i')(Y'_i-\eta_i\Lambda')\right)\right]\\ -->
<!-- & + \mathbb{E}_q\left[\sum_{j=1}^p\left((a_\sigma -1)log(\sigma_j^{-2}) -b_\sigma\sigma_j^{-2}\right)\right]\\ -->
<!-- & + \mathbb{E}_q\left[\frac{1}{2}tr(\eta'\eta) \right]\\ -->
<!-- & + \sum_{j=1,h=1}^{p,q}\mathbb{E}_q\left[0.5log(\phi_{jh})-0.5\Lambda_{jh}^2\phi_{jh}\prod_{k=1}^h \delta_k+0.5\sum_{k=1}^h log(\delta_k) \right]\\ -->
<!-- & + \mathbb{E}_q\left[(\frac{1}{ 2} \gamma -1)log(\phi_{jh}) -\frac{\gamma}{ 2}\phi_{jh}\right]\\ -->
<!-- & + \mathbb{E}_q\left[(a_1 -1)log(\delta_{1}) - \delta_{1}\right]\\ -->
<!-- & + \mathbb{E}_q\left[(a_2 -1)log(\delta_{2}) - \delta_{2}\right]\\ -->
<!-- & + \cdots\\ -->
<!-- & + \mathbb{E}_q\left[(a_2 -1)log(\delta_{q}) - \delta_{q}\right] -->
<!-- \end{align*} -->


<!-- Note that in all cases, expectations are  analytically available. -->

## Updates for the coordinate ascent algorithm

In our context, we want to maximize the ELBO with respect to 
$$\theta = \left\lbrace \left(M^{\Lambda_j}, V^{\Lambda_j}\right)_{1\leq j \leq p}, \left(M^{\eta_i}, V^{\eta_i}\right)_{1\leq i \leq n}, \left(A^{\sigma_j}, B^{\sigma_j}\right)_{1\leq j \leq p}, \left(A^{\phi_{j,h}}, B^{\phi_{j,h}}\right)_{1\leq j \leq p,1\leq h \leq q}, \left(A^{\delta_h}, B^{\delta_h}\right)_{1\leq h \leq q} \right\rbrace.$$
A convenient way to maximize the ELBO in the context of conjugate mean field family is to proceed iteratively via a coordinate ascent algorithm.

Starting from an initial guess $\theta^{(0)}$, we build a sequence $\left(\theta^{(k)}\right)_{k\geq 0}$ using the following procedure:
For any unknown component $Z_u$, the variational parameters are updated by identifying the distribution given by $\mathbb{E}{q_{-u}}\left[f(Z_u,Z_{-u})\right]$, where $Z_{-u}$ denotes all the hidden random variables except $Z_u$, $\mathbb{E}_{q_{-u}}$ denotes the expectation with respect to these random variables, and $f(Z_u,Z_{-u})$ denotes all the terms in the ELBO that depend on $Z_u$ (and potentially $Z_{-u}$). 

Let's illustrate this on the random variable $\sigma^{-2}_j$:

### Updates for $\sigma^{-2}_j$

In the ELBO, we see that
$$
f_{\sigma^{-2}_j}(Z_{\sigma^{-2}_j}, Z_{-\sigma^{-2}_j}) = (a_\sigma -1)\log(\sigma_j^{-2}) -b_\sigma\sigma_j^{-2} + \frac{n}{2} \log \sigma^{-2}_j - \frac{1}{2} \sigma^{-2}_j\left(Y^{j} - \eta \Lambda_j\right)^T\left(Y^{j} - \eta \Lambda_j\right)
$$

Hence, 
$$
\mathbb{E}_{q_{-\sigma^{-2}_j}}[f_{\sigma^{-2}_j}(Z_{\sigma^{-2}_j}, Z_{-\sigma^{-2}_j})] = \left(a_\sigma + \frac{n}{2} - 1\right)\log \sigma^{-2}_j - \left(b_{\sigma} + \frac{1}{2}\mathbb{E}_{q_{\eta,\Lambda_j}}\left[\left(Y^{j} - \eta \Lambda_j\right)^T\left(Y^{j} - \eta \Lambda_j\right)\right] \right)\sigma^{-2}_j + \text{Cst}.
$$
Therefore, one can immediatly recognize the log p.d.f. of a Gamma distribution whose parameters are given by:

* $A^{\sigma_j}= a_\sigma+\frac{n}{2}$; 
* $B^{\sigma_j}_j= b_\sigma + \frac{1}{2}\mathbb{E}_{q_{\eta,\Lambda_j}}\left[\left(Y^{j} - \eta \Lambda_j\right)^T\left(Y^{j} - \eta \Lambda_j\right)\right]$. 

It remains to compute the expectation. First, note that we have:
$$\mathbb{E}\left[ \left(Y^{j} - \eta \Lambda_j\right)^T\left(Y^{j} - \eta \Lambda_j\right) \right] = (Y^{j})^TY^{j} - 2 (Y^{j})^T\mathbb{E}[\eta] \mathbb{E}[\Lambda_j] + \text{Tr}\left(\mathbb{E}[\eta^T\eta]\mathbb{E}\left[\Lambda_j \Lambda_j^T\right] \right).$$

Now, use the fact that (denoting this time $\eta_i$ the $i$-th row of $\eta$):
$$\mathbb{E}[\eta^T\eta] = \sum_{i = 1}^n\mathbb{E}\left[\eta_i\eta_i^T\right] = \sum_{i = 1}^n\left(\mathbb{E}\left[\eta_i\right]\mathbb{E}\left[\eta_i\right]^T + \mathbb{V}\left[\eta_i\right]\right).$$
Suppose we are at iteration $k$ of the algorithm, and therefore that out current guess is $\theta^{(k)}$., we then have that:
\begin{align*}
A^{\sigma_j} &
= a_\sigma+\frac{n}{2}
\\
B^{\sigma_j} &
= b_\sigma + \frac{1}{2}\left[(Y^{j})^TY^{j} - 2 (Y^{j})^TM^\eta M^{\Lambda_j} + \text{Tr}\left( \left(M^{\eta_i}\left(M^{\eta_i} \right)^T + V^{\eta_i}\right) \times \left(M^{\Lambda_j}\left(M^{\Lambda_j}\right)^T + V^{\Lambda_j}\right) \right)\right],
\end{align*}
where all the parameters are the one at the current iteration (we omit the dependance on $(k)$), and $M^\eta$ is the matrix stacking row-wise all the $M^{\eta_i}$.

**Remark on the computation** Note that the computation requires to compute the trace of a matrix product. 
A computational trick here to avoid potentially big matrix product is to notice that when the 2 involved matrices are symmetric (which is the case here), the trace of the matrix product is equal to the sum of all the elements of the matrix given by the element-wise product (also called the Hadamard product). Formally, for symmetric $n\times n$ matrices  $A$ and $B$:
$$\text{Tr}\left({AB}\right) = \sum_{i , j = 1}^n (A\odot B)_{i,j}$$

This trick can lead to critical reduction in computation time (a factor 20 for a 100$\times$ 100 matrix, in `R`).

This is implemented in `R` within  the function 

```{r get_update_VI_Sigma}
get_update_VI_Sigma <- function(Y, params, priors){
  # Useful quantities
  n <- nrow(Y)
  p <- ncol(Y)
  Lambda <- params$Lambda
  Eta <- params$Eta
  Phi <- params$Phi
  Delta <- params$Delta
  # Update A (which could actually be done once for all)
  A <- rep(priors$Sigma$A + n / 2, p)
  # First, compute E[eta'eta]
  E_eta_prime_eta <- apply(Eta$Cov, c(1, 2), sum) + # Sum of variances
    Reduce(f = "+", 
           x = map(1:n, function(i){
             Eta$M[, i] %*% t(Eta$M[, i])
           }))
  # Then, the function to get B^{\sigma_j}
  get_B_sigma_j <- function(j){
    term1 <- sum(0.5 * sum(Y[, j] * Y[, j]))
    term2 <- -sum(Y[, j] * (t(Eta$M) %*% Lambda$M[,j])) # Eta$M is coded in q x n
    # For the last term, we use the element wise product
    term3 <- 0.5 * sum(E_eta_prime_eta * 
                         (Lambda$Cov[,, j] + Lambda$M[, j] %*% t(Lambda$M[, j]))) 
    term1 + term2 + term3
  }
  B <- priors$Sigma$B + map_dbl(1:p, get_B_sigma_j)
  list(A = A, B = B)
}
```

### Updates for $\Lambda_j$

Each term $\Lambda_j$ appears through a quadratic form:

$$-\frac{1}{2}\left(\Lambda_j^{T}\left(\sigma^{-2}_j\eta^{T}\eta + \text{diag}\left(\phi_{j,h}\prod_{\ell = 1}^h\delta_\ell\right)_{1\leq h \leq q}\right)\Lambda_j - 2 \Lambda_j^{T}\sigma^{-2}_j\eta^{T}Y^j\right).$$
Therefore, the CAVI algorithm leads to a Gaussian update, given by:

\begin{align*}
V^{\Lambda_j}&=\left(\frac{A^{\sigma_j}}{B^{\sigma_j}}\sum_{i=1}^n \left(V^{\eta_i}+{M^{\eta_i}}{M^{\eta_i}}'\right)+\text{diag}\left(\frac{A^{\phi_{jh}}}{B^{\phi_{jh}}} \prod_{\ell=1}^h\frac{A^{\delta_{\ell}}}{B^{\delta_{\ell}}} \right) \right)^{-1}\\
M^{\Lambda_j} &= \frac{A^{\sigma_j}}{B^{\sigma_j}} V^{\Lambda_j}  \left(M^{\eta}\right)^TY^{j}.
\end{align*}

The `get_update_VI_Lambda` function implements these equations 
```{r get_update_VI_Lambda}
get_update_VI_Lambda <- function(Y, params){
  # Useful quantities
  n <- nrow(Y)
  p <- ncol(Y)
  Eta <- params$Eta
  Sigma <- params$Sigma
  Phi <- params$Phi
  Delta <- params$Delta
  # First, compute E[eta'eta]
  E_eta_prime_eta <- apply(Eta$Cov, c(1, 2), sum) + # Sum of variances
    Reduce(f = "+", 
           x = map(1:n, function(i){
             Eta$M[, i] %*% t(Eta$M[, i])
           }))
  # Computes V^\Lambda
  V_Lambda <- map(1:p, function(j){
    precision <- Sigma$A[j] / Sigma$B[j] * E_eta_prime_eta + 
      diag(Phi$A[j, ] / Phi$B[j, ] * cumprod(Delta$A) / cumprod(Delta$B))
    variance <- solve(precision)
    return(variance)
  }) %>% 
    abind(along = 3) # Binding in an array of size q x q x p
  M_Lambda <- sapply(1:p, function(j){
    Sigma$A[j] / Sigma$B[j]  * V_Lambda[,, j] %*% Eta$M %*% Y[, j]
  })
  list(M = M_Lambda, Cov = V_Lambda)
}
```

### Updates for $\eta_i$

Each term $\eta_i$ appears through a quadratic form:

$$-\frac{1}{2}\left(\eta_i^{T}\left(\Lambda^{T}\Sigma\Lambda + I_q\right)\eta_i - 2 \eta_i^{T}\Lambda^T\Sigma Y_i\right),$$
where $\Sigma = \text{diag}(\sigma^{-2}_j)_{1\leq j \leq p}$. Then, denote $W$ the $p\times q$ matrix $W = \Sigma^{1/2}\Lambda$. One can notice that the $j$-th row of $W$ is given by $W_j = \sigma_j\Lambda_j$. Then, note that $\Lambda^{T}\Sigma\Lambda = W^TW = \sum_{j = 1}^p W_jW_j^T = \sum_{j = 1}^p\sigma^{-2}_j \Lambda_j\Lambda_j^T$.

Therefore, the CAVI algorithm leads to a Gaussian update, given by:

\begin{align*}
V^{\eta_i}&
= \sum_{j = 1}^p \frac{A^{\sigma_j}}{B^{\sigma_j}} \left(V^{\Lambda_j} + M^{\Lambda_j}\left( M^{\Lambda_j}\right)^T\right)\\
M^{\eta_i} &= V^{\eta_i}\left(M^{\Lambda}\right)^T\text{diag}\left(
\frac{A^{\sigma_j}}{B^{\sigma_j}}
\right)_{1\leq j \leq p}Y_i
\end{align*}

```{r get_update_VI_Eta}
get_update_VI_Eta <- function(Y, params){
  # Useful quantities
  n <- nrow(Y)
  p <- ncol(Y)
  Lambda <- params$Lambda
  Sigma <- params$Sigma
  Phi <- params$Phi
  Delta <- params$Delta
  
  #First, get the common covariance matrix
  common_cov <- lapply(1:p, function(j){
    Sigma$A[j] / Sigma$B[j] * (Lambda$M[,j] %*% t(Lambda$M[,j]) + Lambda$Cov[,,j])
  }) %>% 
    Reduce(f = "+") %>% 
    {. + diag(q)} %>% 
    solve()
  # The common matrix for all means
  mean_matrix <- common_cov %*% Lambda$M %*% diag(Sigma$A/Sigma$B)
  # The means
  M <- sapply(1:n, function(i){
    mean_matrix %*% Y[i, ] 
  })
  Cov <- array(common_cov, dim = c(q, q, n))
  list(M = M, Cov = Cov)
}
```

### Updates for $\phi_{j,h}$

The terms implying $\phi_{j,h}$ are the following:

$$ \left(\frac{\nu}{2} + \frac{1}{2} - 1\right)\log \phi_{j,h} - \left(\frac{\nu}{2} + \Lambda_{j,h}^2\prod_{\ell = 1}^h \delta_\ell\right)\phi_{j,h}.$$
Therefore, the updates of the Gamma distribution parameters are given by:

\begin{align*}
A^{\phi_{j, h}} &= \frac{\nu}{2}+\frac{1}{2}\\ 
B^{\phi_{j, h}} & =\frac{\nu}{2}+\frac{1}{2}\left(\left(M^{\Lambda_{j,h}}\right)^2 + V^{\Lambda_j}_{h,h}\right) \prod_{\ell = 1}^h \frac{A^{\delta_{\ell}}}{B^{\delta_{\ell}}}.
\end{align*}

The corresponding `R` function is implemented.

```{r get_update_VI_Phi}
get_update_VI_Phi <- function(Y, params, priors){
  Lambda <- params$Lambda
  Eta <- params$Eta
  Delta <- params$Delta
  
  # le meme pour tous
  A <- matrix(priors$Phi$A + 0.5, p, q)
  
  cumprod_Delta  = cumprod(Delta$A/Delta$B)
  B <- sapply(1:q, function(h){
    priors$Phi$B + 
      .5 * (Lambda$M[h, ]^2 + Lambda$Cov[h, h, ]) * cumprod_Delta[h]
  })
  list(A = A, B = B)
}
```

### Update for $\delta_h$

The tricky part is that information about $\delta_h$ is supported in the last (from $h$ to $q$) columns of $\Lambda$. Writing things down and recognizing a Gamma distribution, we have:

\begin{align*}
A^{\delta_h} & = \left\lbrace 
\begin{array}{lr}
a_1+\frac{pq}{2} & \text{ if } h = 1\\
a_2 + \frac{p(q + 1 - h)}{2} & \text{ if } h > 1
\end{array} \right.\\
B^{\delta_h} &= 1 + \frac{1}{2}\sum_{\ell = h}^q\left(\left(\prod_{s = 1, s\neq h}^\ell\frac{A^{\delta_s}}{B^{\delta_s}}\right)\left(\sum_{j = 1}^p\left(\left(M^{\Lambda_{j, \ell}}\right)^2 + V^{\Lambda_j}_{\ell, \ell}\right)\frac{A^{\phi_{j,\ell}}}{B^{\phi_{j,\ell}}}\right)\right)
\end{align*}

```{r get_update_VI_Delta}
get_update_VI_Delta <- function(Y, params, priors){
  p <- ncol(Y)
  q <- length(params$Delta$A)
  new_A <- priors$Delta$A + 0.5 * p * (q + 1 - (1:q))
  E_phi_L2 <- sapply(1:q, function(h){
    E_phi_h <- params$Phi$A[, h] / params$Phi$B[, h]
    E_L2 <- params$Lambda$M[h, ]^2 + params$Lambda$Cov[h, h, ]
    sum(E_phi_h * E_L2)
  })
  new_B <- params$Delta$B
  new_B[1] <- priors$Delta$B + .5 * sum(E_phi_L2 * cumprod(new_A / new_B) / (new_A[1] / new_B[1]))
  for(k in 2:q){
    E_delta_all <- cumprod(new_A / new_B)
    E_delta_k <- (new_A[k] / new_B[k])
    new_B[k] <- priors$Delta$B  + .5 * sum(E_phi_L2[k:q] * 
                                             E_delta_all[k:q] / E_delta_k )
  }
  list(A = new_A, B = new_B)
}
```

## Computing the ELBO

The ELBO can be easily computed, as all the involved expectations are explicit. 

We can check the evolution of the ELBO across iterations. To compute the ELBO, one needs to recall that

 * the entropy of a Normal ($M,V$) distribution is equal, up to a constant, to $\frac{\log\vert V \vert}{2}$;
 
 * the entropy of a Gamma ($A, B$) distribution is $A - \log B + \log\Gamma(A)+(1-A) \psi(A)$ with $\psi(\cdot)$ being the  digamma function;
 
```{r}
get_entropy_normal <- function(Cov){
  0.5 * log(det(Cov))
}

get_entropy_gamma <- function(A, B){
  A - log(B) + lgamma(A) + (1 - A) * digamma(A)
}

get_expectation_gamma <- function(A, B){
  A / B
}

get_log_expectation_gamma <- function(A, B){
  digamma(A) -  log(B)
}

get_ELBO <- function(Y, params, priors){
  # Variational entropy term
  variational_entropy <- sum(apply(params$Lambda$Cov, 3, 
                                   get_entropy_normal)) + # Lambda
    sum(apply(params$Eta$Cov, 3, get_entropy_normal)) + # Eta
    sum(get_entropy_gamma(params$Sigma$A, params$Sigma$B)) + # Sigma 
    sum(get_entropy_gamma(params$Delta$A, params$Delta$B)) + # Delta 
    sum(get_entropy_gamma(params$Phi$A, params$Phi$B))  # Phi
  # Usefull expectations
  expectations_log_sigma <- get_log_expectation_gamma(params$Sigma$A, params$Sigma$B)
  expectations_sigma <- get_expectation_gamma(params$Sigma$A, params$Sigma$B)
  expectations_log_delta <- get_log_expectation_gamma(params$Delta$A, params$Delta$B)
  expectations_delta <- get_expectation_gamma(params$Delta$A, params$Delta$B)
  expectations_log_phi <- get_log_expectation_gamma(params$Phi$A, params$Phi$B)
  expectations_phi <- get_expectation_gamma(params$Phi$A, params$Phi$B)
  # Likelihood term
  E_eta_prime_eta <- apply(params$Eta$Cov, c(1, 2), sum) + # Sum of variances
    Reduce(f = "+", 
           x = map(1:n, function(i){
             params$Eta$M[, i] %*% t(params$Eta$M[, i])
           }))
  get_E_quadr_form <- function(j){
    term1 <- sum(0.5 * sum(Y[, j] * Y[, j]))
    term2 <- -sum(Y[, j] * (t(params$Eta$M) %*% params$Lambda$M[,j])) # Eta$M is coded in q x n
    term3 <- 0.5 * sum(E_eta_prime_eta * 
                         (params$Lambda$Cov[,, j] + params$Lambda$M[, j] %*% t(params$Lambda$M[, j]))) 
    term1 + term2 + term3
  }
  likelihood_expectation <- sum(.5 * n * expectations_log_sigma -
                                  expectations_sigma * map_dbl(1:p, get_E_quadr_form))
  # Priors terms
  prior_sigma_expectation <- sum((priors$Sigma$A - 1) * expectations_log_sigma - 
                                   priors$Sigma$B * expectations_sigma)
  prior_phi_expectation <- sum((priors$Phi$A - 1) * expectations_log_phi - 
                                 priors$Phi$B * expectations_phi)
  prior_delta_expectation <- sum((priors$Delta$A - 1) * expectations_log_delta - 
                                   priors$Delta$B * expectations_delta)
  prior_eta_expectation <- -0.5 * sum(map_dbl(1:n, function(i){
    sum(params$Eta$M[,i]^2 + diag(params$Eta$Cov[,, i]))
  }))
  prior_lambda_expectation <- 0.5 * p * sum(cumsum(expectations_log_delta)) +
    0.5 * sum(expectations_log_phi) - 
    0.5 * sum(map_dbl(1:p, function(j){
      sum(cumprod(expectations_delta) * diag(expectations_phi[j, ]) * 
            (diag(params$Lambda$Cov[,, j]) + params$Lambda$M[, j]^2))
    }))
  ELBO <- variational_entropy +
    likelihood_expectation +
    prior_delta_expectation +
    prior_eta_expectation +
    prior_lambda_expectation +
    prior_sigma_expectation +
    prior_phi_expectation
  return(ELBO)
}
```

Therefore, the implementation is now straight forward:

```{r get_CAVI}
get_CAVI <- function(data_, # Y
                     q, # q 
                     n_steps, # Number of steps 
                     params, # initial parameters
                     priors = list(Sigma = list(A = 1, B = 3), 
                                   Phi = list(A = 3/2, B = 3/2),
                                   Delta= list(A = c(2, rep(3, q - 1)), 
                                               B = 1))){
  # Propagation
  # progess_bar <- txtProgressBar(min = 0, max = n_steps)
  current_ELBO <- get_ELBO(data_, params, priors)
  ELBOS <- data.frame(iteration = 0, 
                      ELBO = current_ELBO)
  for(step_ in 1:n_steps){
      params$Lambda <- get_update_VI_Lambda(data_, params)
    # Etas
      params$Eta <- get_update_VI_Eta(data_, params)
    # Sigma
      params$Sigma <- get_update_VI_Sigma(data_, params, priors)
    # Phis
      params$Phi <- get_update_VI_Phi(data_, params, priors)
      # deltas
      params$Delta <- get_update_VI_Delta(data_, params, priors)
      ELBOS <- bind_rows(ELBOS,
                         data.frame(iteration = step_,
                                    ELBO = get_ELBO(data_, params, priors)))
  }
  return(list(ELBOS = ELBOS, params = params))
}
```


# Comparaison with MCMC inference

Relying on Jags, we can get MCMC samples from the posterior distribution of the unknows.

```{r modelString}
modelString = "
model{
#a1 <- 2
#a2 <- 3
a[1] ~ dgamma(2,1)T(2,)
a[2] ~ dgamma(2,1)T(3,)
deminu <- 3/2
a_sigma <- 0.1
b_sigma <- 0.1

#shrinkage prior
for( h in 2:q){delta[h] ~ dgamma(a[2],1)}
              delta[1] ~ dgamma(a[1],1)
for( h in 1:q){ tau[h] <- prod(delta[1:h])} 
for( j in 1:p){
               for( h in 1:q){
                              phi[j,h] ~ dgamma(deminu,deminu)
                              precilambda[j,h] <- phi[j,h]*tau[h]
                              lambda[j,h]~ dnorm(0,precilambda[j,h])
                              }
               }
# N(0,1) prior for eta
for(i in 1:n){
              for (h in 1:q){
                      eta[i,h] ~ dnorm(0,1)
                            }
}
muY <- eta %*% t(lambda)
# finally Y
for(j in 1:p){ preciE[j] ~ dgamma(a_sigma,b_sigma)
              for (i in 1:n){
                      Y[i,j] ~ dnorm(muY[i,j],preciE[j])
                            }
}

}"
```


```{r jags_inference, eval = FALSE}
data_for_JAGS <- list(
  n = dim(Y)[1],
  p = dim(Y)[2],
  q = 6,
  Y = Y
)
n.adapt = 1000
burnin = 1000
n.iter = burnin * 2
thin = 1
jm <- jags.model(
  file = textConnection(modelString),
  data = data_for_JAGS,
  n.chains = 3 ,
  n.adapt = n.adapt
)
update(jm, burnin)
jsamples <- coda.samples(
  model = jm,
  variable.names = c("tau","a"),
  n.iter = n.iter,
  thin = thin
)
```


We transform the MCMC coda file into a regular tibble to visualize posterior pdfs and variational ones.

```{r MCMCvisualization}
MCMC_results <- ggs(jsamples)
ggs_traceplot(MCMC_results)
ggs_density(MCMC_results %>%
              filter(Parameter %in% c("a[1]","a[2]")))
ggs_density(MCMC_results,family=c("tau"))
ggs_caterpillar(MCMC_results,family = "tau") +
  scale_x_continuous(trans = "log10")
```

We can now visualize the estimated posterior distributions for all parameters:


