---
title: "About Variational Bayes for PPCA data"
author: "Eric PARENT & Achille Thin & Pierre Gloaguen"
date: "the 3rd of January 2023 , UMR MIA-Paris, AgroParisTech, INRA, Université Paris-Saclay"
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_document:
    code_folding: hide
    df_print: paged
    self_contained: yes
    number_sections: yes
    toc: yes
    theme: journal
    toc_float: yes
    toc_depth: 2
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE, cache = FALSE}

# knitr::opts_chunk$set(
# 	echo = TRUE,
# 	message = FALSE,
# 	warning = FALSE,
# 	cache = FALSE
# )
knitr::opts_chunk$set(
	eval = FALSE,
	message = FALSE,
	warning = FALSE,
	cache = FALSE,
	include = TRUE
)
```

# R codes and librairies {-}

All `R` codes can be unfolded, for instance, unfold the following chunk to see which libairies are used in this document:

```{r librairies}
rm(list=ls())
library(tidyverse) # For data manipulation and visualization
library(rjags)  # For MCMC sampling
library(ggmcmc) # To format jags output
library(extraDistr) # For the non standart t distribution
```


# A not-so-simple shrinkage PPCA model

Bhattacharya and Dunson consider the following linear normal model with a gamma process to shrink the variances of the columns of the latent matrix $\lambda$. Relying on this setting, we include a fixed effect (due to $f=1:F_x$ forcing conditions from known explanatary variates with intensity $X_{f,i}$ on site $i$ interveining with coefficient $\beta_{j,f}$ onto species $j=1:p$) in addition to the random ones (governed by the unknown sparse loading matrix $\Lambda$):
\begin{align*}
Y_i &
\overset{ind.}{\sim} \mathcal{N}_p(\beta X_i+\Lambda\eta_i, \Sigma), &
1\leq i \leq n 
\\
\Sigma &= 
\text{diag}(\sigma_j^{2}), &
1\leq j \leq p 
\\
\sigma_j^{-2}|(a_{\sigma},b_{\sigma}) &
\sim \Gamma(a_{\sigma},b_{\sigma}) &
1\leq j \leq p
\\
\Lambda_{jh} &
\sim \mathcal{N}_{q}(0,\,\phi_{jh}^{-1}\tau_{h}^{-1}), &
1\leq j \leq p, 1 \leq h \leq {q}
\\
\phi_{jh} &
\sim \Gamma(\frac{\nu}{2},\frac{\nu}{2}), &
1\leq j \leq p, 1 \leq h \leq {q}\\
\tau_{h} &
= \prod_{l=1}^h \delta_l &
1\leq h \leq q
\\ 
\delta_1 &
\sim \Gamma(a_1,1) &
\\
\delta_h &
\sim \Gamma(a_2,1) &
\forall l \geq 2
\end{align*}

For computational convenience, we suggest to consider the following independent scaling prior for $\beta$:

\begin{align*}
\beta_{j,f} & \sim \mathcal{N}(0,c_f \times \sigma^2_j) &
1\leq j \leq p ,
1\leq f \leq F_x
\end{align*}

with $c_f$ a supposedly known scaling constant.

Recall that -- up to constant terms-- the logdensity of such a Normal multivariate distribution is:
$$\log[Y|\Lambda,\Sigma,\beta] = -\frac{n}{2}\left(\log\vert C\vert +\text{Tr}(C^{-1}\hat{S}_{\tilde{Y}})\right)$$
with

$$ C= \Lambda\Lambda'+\Sigma\\
\hat{S}_{\tilde{Y}}=\frac{1}{n}\sum_{i=1}^n (Y_i-\beta X_i)'(Y_i-\beta X_i)$$

In the case $q<<p$, because $\Sigma$ is diagonal, it is easier to compute the inverse of the variance-covariance matrix by $$C^{-1}=\Sigma^{-1}-\Sigma^{-1}\Lambda(I+ \Lambda' \Sigma^{-1}\Lambda)^{-1}\Lambda'\Sigma^{-1}$$


Some example data are generated in the following R code chunk.

```{r generation}
source("utils_PPCA_generating_data_fixed.R") # For true values
Y <- read.table("data_PPCA.txt", sep = ";") %>%
  as.matrix()
X <- read.table("fixed_PPCA.txt", sep = ";") %>%
  as.matrix()
```

In the code, $Y$ is a known $n\times p$ matrix, $X$ is a known $n\times F_x$ matrix while the unknown $\Lambda$ will be a $p \times q$ matrix, $\beta$ will be a $p \times F_x$ matrix and $\eta$ a $n\times q$ matrix.

# Variational approach

## Variational family

We denote $Z$ the unknown with components $(\tau,\phi,\eta,\Lambda,\Sigma, \beta)$ and choose an independent family $q(Z) = \left(\prod_u q_{z_u}(Z_u) \right)$ to approximate $[Z|Y]$ such that :
\begin{align*}
\delta_h &
\sim q_{\delta_h}(\cdot)= \mathcal{G}(A^{\delta}_h,B^{\delta}_h), &
1\leq h \leq q 
\\
\phi_{jh} &
\sim q_{\phi_{jh}}(\cdot)= \mathcal{G}(A^{\phi}_{jh},B^{\phi}_{jh}),  &
1\leq j \leq p, 1\leq h \leq q
\\
\sigma^{-2}_{j} &
\sim q_{\sigma^{-2}_{j}}(\cdot) = \mathcal{G}(A^{\sigma}_j,B^{\sigma}_j), &
1\leq j \leq p 
\\
\eta_i &
\sim q_{\eta_i}(\cdot)= \mathcal{N}(M_{\eta_i}, V^{\eta_i}), & 
1\leq i \leq n
\\
\Lambda_j &
\sim q_{\Lambda_j }(\cdot)= \mathcal{N}(M_{\lambda_j},V^{\Lambda_j}), &
1\leq j \leq p 
\\
\beta_j &
\sim q_{\beta_j }(\cdot)= \mathcal{N}(M_{\beta_j},V^{\beta_j}), &
1\leq j \leq p, 
\end{align*}
where $\eta_i$ denotes the $i$-th row of yhe $n\times q$ matrix $\eta$,  $\Lambda_j$ denotes the $j$-th row of the $p\times q$ matrix $\Lambda$,and $\beta_j$ denotes the $j$-th row of the $p\times F_x$ matrix $\beta$. In the following, all vectors are treated as column vectors.

## Evidence lower bound

To update parameters, we rely on the minimisation of the Kullback Liebler divergence
$KL= \mathbb{E}_q\left[\log\frac{q(Z)}{[Z|Y]}\right]$.
Equivalently we want to maximize the ELBO

$$\text{ELBO}= \mathbb{E}_q\left[\log[Y,\tau,\phi,\eta,\Lambda,\sigma^{-2},\beta]\right]-\mathbb{E}_q\left[\log q(\tau,\phi,\eta,\Lambda,\sigma^{-2},\beta)\right]$$

Due to conditionnal independence the criterion can decomposed as 

\begin{align*}
\text{ELBO}=&\mathbb{E}_q\left[\log\left([Y|\eta,\Lambda,\Sigma,\beta]\right)\right]\\
& + \mathbb{E}_q\left[\log\left([\beta| \Sigma]\right)\right]\\
& + \mathbb{E}_q\left[\log\left([\Sigma]\right)\right]\\
& + \mathbb{E}_q\left[\log\left([\eta]\right) \right]\\
& + \mathbb{E}_q\left[\log\left([\Lambda|\delta,\phi]\right)\right]\\
& + \mathbb{E}_q\left[\log\left([\phi]\right)\right]\\
& + \mathbb{E}_q\left[\log\left([\delta]\right)\right]\\
& -\mathbb{E}_q\left[\log q(\Lambda,\Sigma,\eta,\phi,\delta, \beta)\right]
\end{align*}

## Terms to integrate

Up to constant terms:
\begin{align*}
\log[Y|\Lambda,\Sigma,\eta,\beta]  &
= \frac{n}{2}\sum_{j = 1}^p \log \sigma^{-2}_j - \frac{1}{2}\sum_{j = 1}^p \sigma^{-2}_j\left(Y^{j} - \eta \Lambda_j-X\beta_j\right)^T\left(Y^{j} - \eta \Lambda_j-X\beta_j\right)
\\
&= -\frac{n}{2}\left(\log\vert \Sigma^{-1}\vert +\text{Tr}(\Sigma^{-1}\hat{S}_{Y,\eta,\Lambda})\right),
\end{align*}
with

$$ \hat{S}_{Y,\eta,\Lambda}=\frac{1}{n}\sum_{i=1}^n (Y_i-\Lambda\eta_i'-\beta X_i')(Y'_i-\eta_i\Lambda'- X_i\beta')\\
= \frac{1}{n} (Y'-\eta\Lambda'-X\beta')(Y-\Lambda\eta'-\beta X').$$


Continuing (still up to constant terms), we have for $\eta$:
$$\log[\eta] = -\frac{1}{2}\sum_{i = 1}^n \eta_i^T\eta_i = -\frac{1}{2}\text{Tr}\left(\eta'\eta\right).$$
Due to prior normal assuptions for $\beta$ : 
$$\log\left([\beta|\Sigma]\right)= 0.5\times F_x\sum_j^p log(\sigma_j^{-2}) -0.5\times p\sum_f^{F_x} log(c_f)\\-0.5 \sum_{j,f}^{p,F_x} c_f^{-1} \sigma_j^{-2}\beta_{jf}^2$$
Returning to vectors ($\beta_j$ is the $j$th row of matrix $\beta$):

$$\log\left([\beta_j|\Sigma]\right)= 0.5\times F_x \times log(\sigma_j^{-2}) -0.5\times \sigma_j^{-2} \times Tr(diag(c_f^{-1})\beta_j \beta_j^T)$$

Since $\Sigma$ is diagonal:

$$\log[\Sigma^{-1}] = \sum_{j=1}^p\left((a_\sigma -1)\log(\sigma_j^{-2}) -b_\sigma\sigma_j^{-2}\right)$$

Due to normal assumptions for $\Lambda_j$

$$\log\left([\Lambda_{j}|\tau_{h},\phi_{j,h}]\right)= \frac{1}{2}\sum_{h = 1}^q \left(\log \phi_{j,h} + \sum_{\ell = 1}^h\log \delta_\ell\right) - \frac{1}{2}\text{Tr}\left(\text{diag}\left(\phi_{j,h}\prod_{\ell = 1}^h\delta_\ell\right)\Lambda_{j}\Lambda_{j}^T \right).$$
Remains the terms with the prior for $\phi_{j,h}\;(1\leq j \leq p, 1\leq h \leq q)$:

$$\log[\phi_{jh}] = \left(\frac{\nu}{ 2} - 1\right)\log(\phi_{j,h}) -\frac{\nu}{ 2}\phi_{jh},$$
and $\delta_h\; (1\leq h \leq q)$

$$
\log[\delta_{h}] = \left\lbrace 
\begin{array}{lr}
(a_1 -1)\log(\delta_{1}) -b_{1} \delta_{1} & \text{ if } h = 1,\\
(a_2 -1)\log(\delta_{2}) -b_{2} \delta_{2} & \text{ else.}
\end{array}
\right.
$$


<!-- Wrapping up all the previous log density terms -->


<!-- \begin{align*} -->
<!-- \text{ELBO}=& -\mathbb{E}_q\left[\log q(\Lambda,\Sigma,\eta,\phi,\delta)\right]\\ -->
<!-- & +\mathbb{E}_q\left[\frac{n}{2}\left(\sum_{j=1}^p \log\sigma_j^{-2} \right)-0.5\sum_{i=1}^n tr\left(\Sigma^{-1} (Y_i-\Lambda\eta_i')(Y'_i-\eta_i\Lambda')\right)\right]\\ -->
<!-- & + \mathbb{E}_q\left[\sum_{j=1}^p\left((a_\sigma -1)log(\sigma_j^{-2}) -b_\sigma\sigma_j^{-2}\right)\right]\\ -->
<!-- & + \mathbb{E}_q\left[\frac{1}{2}tr(\eta'\eta) \right]\\ -->
<!-- & + \sum_{j=1,h=1}^{p,q}\mathbb{E}_q\left[0.5log(\phi_{jh})-0.5\Lambda_{jh}^2\phi_{jh}\prod_{k=1}^h \delta_k+0.5\sum_{k=1}^h log(\delta_k) \right]\\ -->
<!-- & + \mathbb{E}_q\left[(\frac{1}{ 2} \gamma -1)log(\phi_{jh}) -\frac{\gamma}{ 2}\phi_{jh}\right]\\ -->
<!-- & + \mathbb{E}_q\left[(a_1 -1)log(\delta_{1}) - \delta_{1}\right]\\ -->
<!-- & + \mathbb{E}_q\left[(a_2 -1)log(\delta_{2}) - \delta_{2}\right]\\ -->
<!-- & + \cdots\\ -->
<!-- & + \mathbb{E}_q\left[(a_2 -1)log(\delta_{q}) - \delta_{q}\right] -->
<!-- \end{align*} -->


<!-- Note that in all cases, expectations are  analytically available. -->

## Updates for the coordinate ascent algorithm

In our context, we want to maximize the ELBO with respect to 
$$\theta = \left\lbrace \left(M^{\beta_j}, V^{\beta_j}\right)_{1\leq j \leq p}, \left(M^{\Lambda_j}, V^{\Lambda_j}\right)_{1\leq j \leq p}, \left(M^{\eta_i}, V^{\eta_i}\right)_{1\leq i \leq n}, \left(A^{\sigma_j}, B^{\sigma_j}\right)_{1\leq j \leq p}, \left(A^{\phi_{j,h}}, B^{\phi_{j,h}}\right)_{1\leq j \leq p,1\leq h \leq q}, \left(A^{\delta_h}, B^{\delta_h}\right)_{1\leq h \leq q} \right\rbrace.$$
A convenient way to maximize the ELBO in the context of conjugate mean field family is to proceed iteratively via a coordinate ascent algorithm.

Starting from an initial guess $\theta^{(0)}$, we build a sequence $\left(\theta^{(k)}\right)_{k\geq 0}$ using the following procedure:
For any unknown component $Z_u$, the variational parameters are updated by identifying the distribution given by $\mathbb{E}{q_{-u}}\left[f(Z_u,Z_{-u})\right]$, where $Z_{-u}$ denotes all the hidden random variables except $Z_u$, $\mathbb{E}_{q_{-u}}$ denotes the expectation with respect to these random variables, and $f(Z_u,Z_{-u})$ denotes all the terms in the ELBO that depend on $Z_u$ (and potentially $Z_{-u}$). 

Let's illustrate this on the random variable $\sigma^{-2}_j$:

### Updates for $\sigma^{-2}_j$



In the ELBO, we see that
$$
f_{\sigma^{-2}_j}(Z_{\sigma^{-2}_j}, Z_{-\sigma^{-2}_j}) = (a_\sigma -1)\log(\sigma_j^{-2}) -b_\sigma\sigma_j^{-2} + \frac{n+F_x}{2} \log \sigma^{-2}_j - \frac{1}{2} \sigma^{-2}_j\left(Y^{j} - \eta \Lambda_j\right)^T\left(Y^{j} - \eta \Lambda_j\right)
$$

Hence, 
$$
\mathbb{E}_{q_{-\sigma^{-2}_j}}[f_{\sigma^{-2}_j}(Z_{\sigma^{-2}_j}, Z_{-\sigma^{-2}_j})] = \left(a_\sigma + \frac{n}{2} - 1\right)\log \sigma^{-2}_j - \left(b_{\sigma} + \frac{1}{2}\mathbb{E}_{q_{\eta,\Lambda_j}}\left[\left(Y^{j} - \eta \Lambda_j-X\beta_j\right)^T\left(Y^{j} - \eta \Lambda_j-X\beta_j\right)\right] \right)\sigma^{-2}_j + \text{Cst}.
$$
Therefore, one can immediatly recognize the log p.d.f. of a Gamma distribution whose parameters are given by:

* $A^{\sigma_j}= a_\sigma+\frac{n}{2}$; 
* $B^{\sigma_j}_j= b_\sigma + \frac{1}{2}\mathbb{E}_{q_{\eta,\Lambda_j}}\left[\left(Y^{j} - \eta \Lambda_j-X\beta_j\right)^T\left(Y^{j} - \eta \Lambda_j-X\beta_j\right)\right]$. 

It remains to compute the expectation. First, note that we have results for $X=0$ since the term :
$$\mathbb{E}\left[ \left(Y^{j} - \eta \Lambda_j\right)^T\left(Y^{j} - \eta \Lambda_j\right) \right] = (Y^{j})^TY^{j} - 2 (Y^{j})^T\mathbb{E}[\eta] \mathbb{E}[\Lambda_j] + \text{Tr}\left(\mathbb{E}[\eta^T\eta]\mathbb{E}\left[\Lambda_j \Lambda_j^T\right] \right).$$ 
has been previously obtained. For clarity, the previous function has been renamed *get_update_VI_Sigma_without_fixed_effects*.  It remains to add a quadratic component in $\beta_j$ and a rectangular one, linear in $X\beta_j$

This is implemented in `R` within  the following function (in which $X'X$ has been precomputed for computer time saving):

```{r get_update_VI_Sigma}
get_update_VI_Sigma <- function(Y, params, priors,
                                X = 0, XprimeX = 0){
  # Useful quantities
  n <- nrow(Y)
  p <- ncol(Y)
  F_x <- ncol(X)
  Lambda <- params$Lambda
  Eta <- params$Eta
  Beta <- params$Beta
  # Posterior variationel de A
  update_without_X <- get_update_VI_Sigma_without_fixed_effects(Y, params, priors)
  if(all(X == 0)){ # Case when no covariates
    return(list(A = update_without_X$A, B = update_without_X$B))
  }
  else{
    A <- update_without_X$A + F_x * 0.5
    get_updade_term_B_sigma_j <- function(j){
      term1 <- 0.5 * sum((XprimeX + priors$Beta$C) * (Beta$Cov[,,j] + Beta$M[,j] %*% t(Beta$M[,j])))
      term2 <- -sum((Y[, j] -  t(Eta$M) %*% Lambda$M[,j]) * (X %*% Beta$M[,j])) # Eta$M is coded in q x n
      term1 + term2
    }
    B <- update_without_X$B + map_dbl(1:p, get_updade_term_B_sigma_j) 
  }
  return(list(A = A, B = B))
}

```

### Updates for $\Lambda_j$

Each term $\Lambda_j$ appears through a quadratic form:

$$-\frac{1}{2}\left(\Lambda_j^{T}\left(\sigma^{-2}_j\eta^{T}\eta + \text{diag}\left(\phi_{j,h}\prod_{\ell = 1}^h\delta_\ell\right)_{1\leq h \leq q}\right)\Lambda_j - 2 \Lambda_j^{T}\sigma^{-2}_j\eta^{T}(Y^j-X\beta_j)\right).$$
Therefore, as there is no quadratic terms in $\beta$ the CAVI algorithm leads to a Gaussian update, relying on the same function as the one without fixed effects where $Y_j$ is replaced by $Y_j-X\mathbb{E}(\beta_j)$.


### Updates for $\eta_i$

Each term $\eta_i$ appears through a quadratic form in $\eta$ but a linear one in $Y$:

$$-\frac{1}{2}\left(\eta_i^{T}\left(\Lambda^{T}\Sigma\Lambda + I_q\right)\eta_i - 2 \eta_i^{T}\Lambda^T\Sigma (Y_i-\beta X_i)\right),$$

Therefore, the CAVI algorithm leads to a Gaussian update, relying on the same function as the one without fixed effects where $Y_j$ is replaced by $Y_j-X\mathbb{E}(\beta_j)$.

\begin{align*}
V^{\eta_i}&
= \sum_{j = 1}^p \frac{A^{\sigma_j}}{B^{\sigma_j}} \left(V^{\Lambda_j} + M^{\Lambda_j}\left( M^{\Lambda_j}\right)^T\right)\\
M^{\eta_i} &= V^{\eta_i}\left(M^{\Lambda}\right)^T\text{diag}\left(
\frac{A^{\sigma_j}}{B^{\sigma_j}}
\right)_{1\leq j \leq p}Y_i
\end{align*}

### Updates for $\phi_{j,h}$

The terms implying $\phi_{j,h}$ are the following:

$$ \left(\frac{\nu}{2} + \frac{1}{2} - 1\right)\log \phi_{j,h} - \left(\frac{\nu}{2} + \Lambda_{j,h}^2\prod_{\ell = 1}^h \delta_\ell\right)\phi_{j,h}.$$
Therefore, the updates of the Gamma distribution parameters are given by:

\begin{align*}
A^{\phi_{j, h}} &= \frac{\nu}{2}+\frac{1}{2}\\ 
B^{\phi_{j, h}} & =\frac{\nu}{2}+\frac{1}{2}\left(\left(M^{\Lambda_{j,h}}\right)^2 + V^{\Lambda_j}_{h,h}\right) \prod_{\ell = 1}^h \frac{A^{\delta_{\ell}}}{B^{\delta_{\ell}}}.
\end{align*}

With regards to absence of $Y$ in these updating equations, the updating sub routine  without fixed effect for $phi$ can be kept without any change.

### Update for $\delta_h$

One can rely of the updating subroutine of $\delta_h$ from the model without fixed effect that doe not depend on $Y$.


### Updates for $\beta_j$

Sine $\beta$ plays a role similar to $\Lambda$ in the updating algorithm, it is easily found that $\beta_j$ has normal variational distribution obtained by applying conjugate properties. This is derived from the structure of the terms involving $\beta$ in 
$$ \frac{-\sigma^{-2}_j }{2}\left((Y^{j} - \eta \Lambda_j-X\beta_j\right)^T\left(Y^{j} - \eta \Lambda_j-X\beta_j)+Tr(diag(c_f)\beta_j\beta_j')\right) $$

```{r get_update_VI_Beta}
get_update_VI_Beta <- function(Y, params, priors, X, XprimeX){
  n <- nrow(Y)
  p <- ncol(Y)
  Sigma <- params$Sigma
  # Calcul des variances
  V_Beta <- lapply(1:p, function(j){
    precision <- Sigma$A[j] / Sigma$B[j] * (XprimeX + priors$Beta$C)
    variance <- solve(precision)
    return(variance)
  }) %>% 
    abind(along = 3) # Mise au format array
  M_Beta <- sapply(1:p, function(j){
    Sigma$A[j] / Sigma$B[j]  * V_Beta[,, j] %*% t(X) %*% (Y[, j] - t(params$Eta$M) %*% params$Lambda$M[, j])
  })
  list(M = M_Beta, Cov = V_Beta)
}

```


**To be Continued**

## Computing the ELBO

The ELBO can be easily computed, as all the involved expectations are explicit. 

We can check the evolution of the ELBO across iterations. To compute the ELBO, one needs to recall that

 * the entropy of a Normal ($M,V$) distribution is equal, up to a constant, to $\frac{\log\vert V \vert}{2}$;
 
 * the entropy of a Gamma ($A, B$) distribution is $A - \log B + \log\Gamma(A)+(1-A) \psi(A)$ with $\psi(\cdot)$ being the  digamma function;
 
```{r}
get_ELBO <- function(Y, params, priors, X = 0, XprimeX = 0){
  # Variational entropy term
  variational_entropy <- sum(apply(params$Lambda$Cov, 3, 
                                   get_entropy_normal)) + # Lambda
    sum(apply(params$Eta$Cov, 3, get_entropy_normal)) + # Eta
    sum(get_entropy_gamma(params$Sigma$A, params$Sigma$B)) + # Sigma 
    sum(get_entropy_gamma(params$Delta$A, params$Delta$B)) + # Delta 
    sum(get_entropy_gamma(params$Phi$A, params$Phi$B)) # Phi
  if(all(X != 0)){ # Case with covariates X with parameter beta
    variational_entropy <- variational_entropy +
      sum(apply(params$Beta$Cov, 3, get_entropy_normal))  # Beta
  }
  # Usefull expectations
  expectations_log_sigma <- get_log_expectation_gamma(params$Sigma$A, params$Sigma$B)
  expectations_sigma <- get_expectation_gamma(params$Sigma$A, params$Sigma$B)
  expectations_log_delta <- get_log_expectation_gamma(params$Delta$A, params$Delta$B)
  expectations_delta <- get_expectation_gamma(params$Delta$A, params$Delta$B)
  expectations_log_phi <- get_log_expectation_gamma(params$Phi$A, params$Phi$B)
  expectations_phi <- get_expectation_gamma(params$Phi$A, params$Phi$B)
  # Likelihood term
  E_eta_prime_eta <- apply(params$Eta$Cov, c(1, 2), sum) + # Sum of variances
    Reduce(f = "+", 
           x = lapply(1:n, function(i){
             params$Eta$M[, i] %*% t(params$Eta$M[, i])
           }))
  get_E_quadr_form <- function(j){
    term1 <- sum(0.5 * sum(Y[, j] * Y[, j]))
    term2 <- - sum(Y[, j] * (t(params$Eta$M) %*% params$Lambda$M[,j])) # Eta$M is coded in q x n
    term3 <- 0.5 * sum(E_eta_prime_eta * 
                         (params$Lambda$Cov[,, j] + params$Lambda$M[, j] %*% t(params$Lambda$M[, j]))) 
    term2bis <- 0
    term3bis <- 0
    term4 <- 0
    if(all(X != 0)){ # Case with covariates X with parameter beta
      term2bis <- -sum(Y[, j] * (X %*% params$Beta$M[,j]) ) # Eta$M is coded in q x n
      term3bis <- 0.5 * sum(XprimeX * 
                              (params$Beta$Cov[,, j] + params$Beta$M[, j] %*% t(params$Beta$M[, j]))) 
      term4 <- sum((X %*% params$Beta$M[,j])*t(params$Lambda$M[,j]%*%params$Eta$M))
    }
    
    term1 + term2 + term3 +term2bis +term3bis+term4
  }
  likelihood_expectation <- sum(.5 * n * expectations_log_sigma -
                                  expectations_sigma * map_dbl(1:p, get_E_quadr_form))
  # Priors terms
  prior_sigma_expectation <- sum((priors$Sigma$A - 1) * expectations_log_sigma - 
                                   priors$Sigma$B * expectations_sigma)
  prior_phi_expectation <- sum((priors$Phi$A - 1) * expectations_log_phi - 
                                 priors$Phi$B * expectations_phi)
  prior_delta_expectation <- sum((priors$Delta$A - 1) * expectations_log_delta - 
                                   priors$Delta$B * expectations_delta)
  prior_eta_expectation <- -0.5 * sum(map_dbl(1:n, function(i){
    sum(params$Eta$M[,i]^2 + diag(params$Eta$Cov[,, i]))
  }))
  prior_lambda_expectation <- 0.5 * p * sum(cumsum(expectations_log_delta)) +
    0.5 * sum(expectations_log_phi) - 
    0.5 * sum(map_dbl(1:p, function(j){
      sum(cumprod(expectations_delta) * diag(expectations_phi[j, ]) * 
            (diag(params$Lambda$Cov[,, j]) + params$Lambda$M[, j]^2))
    }))
  prior_beta_expectation <- 0
  if(all(X != 0)){ 
    prior_beta_expectation <- 0.5 * F_x * sum(expectations_log_sigma) -
      0.5 * sum(map_dbl(1:p, function(j){
        expectations_sigma[j] *sum(diag(priors$Beta$C) * 
                                     (params$Beta$Cov[,, j] + diag(params$Beta$M[, j]^2)))
      }))
  }
  ELBO <- variational_entropy +
    likelihood_expectation +
    prior_delta_expectation +
    prior_eta_expectation +
    prior_lambda_expectation +
    prior_sigma_expectation +
    prior_phi_expectation +
    prior_beta_expectation
  return(ELBO)
}


```

Therefore, the implementation is now straightforward:

```{r get_CAVI}
get_CAVI <- function(data_, q, n_steps, 
                     X = NULL, 
                     set_seed = FALSE,
                     params = NULL,
                     priors = NULL,
                     updates = c(Lambda = TRUE, Sigma = TRUE,
                                 Eta = TRUE, Delta = TRUE, Phi = TRUE, 
                                 Beta = TRUE),
                     debug = FALSE,
                     get_ELBO_freq = 1){
  if(is.null(priors)){
    stop("We are bayesian guys, please specify a prior")
  }
  p <- ncol(data_); n <- nrow(data_); 
  if(is.null(X)){
    updates["Beta"] <- FALSE
    X <- matrix(0, nrow = nrow(data_), ncol = 1)
    priors$Beta = list(M = rep(0, ncol(X)),
                       C = rep(0.01, ncol(X)))
  }
  F_x <- ncol(X)
  XprimeX <- t(X) %*% X
  if(set_seed){
    set.seed(set_seed)
  }
  if(is.null(params)){
    params <- list(Lambda = list(M = matrix(rnorm(q * p), q, p),
                                 Cov = array(diag(1, q), dim = c(q, q, p))),
                   Beta = list(M = matrix(rnorm(F_x * p),
                                          nrow = F_x, ncol = p),
                               Cov = array(diag(1, F_x), 
                                           dim = c(F_x, F_x, p))),
                   Eta = list(M = matrix(rnorm(q * n), q, n),
                              Cov = array(diag(1, q), dim = c(q, q, n))),
                   Sigma = list(A = runif(p, 1, 3),
                                B = runif(p, 1, 3)),
                   Delta = list(A = runif(q, 2, 5) ,
                                B = rep(1, q)),
                   Phi = list(A = matrix(3/2, p, q),
                              B = matrix(3/2, p, q)))
  }
  
  # Propagation
  # progess_bar <- txtProgressBar(min = 0, max = n_steps)
  current_ELBO <- get_ELBO(Y = data_, params = params, priors = priors, 
                           X = X, XprimeX = XprimeX)
  ELBOS <- data.frame(iteration = 0, 
                      ELBO = current_ELBO)
  # print(X)
  # print(params$Beta$M)
  for(step_ in 1:n_steps){
    # Lambdas
    if(updates["Lambda"]){
      params$Lambda <- get_update_VI_Lambda(data_ - X %*% params$Beta$M, params)
      if(debug){
        new_ELBO <- get_ELBO(Y = data_, params = params, priors = priors, 
                             X = X, XprimeX = XprimeX)
        if(new_ELBO < current_ELBO){
          print("Problem at iteration", step_, "after updating Lambda")
        }
        current_ELBO <- new_ELBO
      }
    }
    # Etas
    if(updates["Eta"]){
      params$Eta <- get_update_VI_Eta(data_ - X %*% params$Beta$M, params)
      if(debug){
        new_ELBO <- get_ELBO(Y = data_, params = params, priors = priors, 
                             X = X, XprimeX = XprimeX)
        if(new_ELBO < current_ELBO){
          print("Problem at iteration", step_, "after updating Eta")
        }
        current_ELBO <- new_ELBO
      }
    }
    # Sigma
    if(updates["Sigma"]){
      params$Sigma <- get_update_VI_Sigma(data_, params, priors,
                                          X = X, XprimeX = XprimeX)
      if(debug){
        new_ELBO <- get_ELBO(Y = data_, params = params, priors = priors, 
                             X = X, XprimeX = XprimeX)
        if(new_ELBO < current_ELBO){
          print("Problem at iteration", step_, "after updating Sigma")
        }
        current_ELBO <- new_ELBO
      }
    }
    if(updates["Beta"]){
      params$Beta <- get_update_VI_Beta(Y = data_, params, priors, 
                                        X = X, XprimeX = XprimeX)
      if(debug){
        new_ELBO <- get_ELBO(Y = data_, params = params, priors = priors, 
                             X = X, XprimeX = XprimeX)
        if(new_ELBO < current_ELBO){
          print("Problem at iteration", step_, "after updating Beta")
        }
        current_ELBO <- new_ELBO
      }
    }
    # Phis
    if(updates["Phi"]){
      params$Phi <- get_update_VI_Phi(params, priors)
      if(debug){
        new_ELBO <- get_ELBO(Y = data_, params = params, priors = priors, 
                             X = X, XprimeX = XprimeX)
        if(new_ELBO < current_ELBO){
          print("Problem at iteration", step_, "after updating Phi")
        }
        current_ELBO <- new_ELBO
      }
    }
    if(updates["Delta"]){
      # deltas
      params$Delta <- get_update_VI_Delta(params, priors)
      if(debug){
        new_ELBO <- get_ELBO(Y = data_, params = params, priors = priors, 
                             X = X, XprimeX = XprimeX)
        if(new_ELBO < current_ELBO){
          print(paste("Problem at iteration", step_, "after updating Delta"))
        }
        current_ELBO <- new_ELBO
      }
    }
    if((n_steps %% get_ELBO_freq) == 0){
      ELBOS <- bind_rows(ELBOS,
                         data.frame(iteration = step_,
                                    ELBO = get_ELBO(Y = data_, params = params, priors = priors, 
                                                    X = X, XprimeX = XprimeX)))
    }
  }
  return(list(ELBOS = ELBOS, params = params))
}

```


### And run!
```{r}
source("script_PPCA_VI_implementation_fixed.R")
```


# Comparaison with MCMC inference

Relying on Jags, we can get MCMC samples from the posterior distribution of the unknows.

```{r modelString}
modelString = "
model{
#a1 <- 2
#a2 <- 3
a[1] ~ dgamma(2,1)T(2,)
a[2] ~ dgamma(2,1)T(3,)
deminu <- 3/2
a_sigma <- 0.1
b_sigma <- 0.1
precibeta <- 0.01

#shrinkage prior
for( h in 2:q){delta[h] ~ dgamma(a[2],1)}
              delta[1] ~ dgamma(a[1],1)
for( h in 1:q){ tau[h] <- prod(delta[1:h])} 
for( j in 1:p){
               for( h in 1:q){
                              phi[j,h] ~ dgamma(deminu,deminu)
                              precilambda[j,h] <- phi[j,h]*tau[h]
                              lambda[j,h]~ dnorm(0,precilambda[j,h])
                              }
               }
# N(0,1) prior for eta
for(i in 1:n){
              for (h in 1:q){
                      eta[i,h] ~ dnorm(0,1)
                            }
}
for( f in 1:F_x){
               for( j in 1:p){
               beta[j,f]~ dnorm(0,precibeta)
               }
}
muY <- eta %*% t(lambda)+X %*% t(beta)

# finally Y
for(j in 1:p){ preciE[j] ~ dgamma(a_sigma,b_sigma)
              for (i in 1:n){
                      Y[i,j] ~ dnorm(muY[i,j],preciE[j])
                            }
}

}"
```


```{r jags_inference, eval = FALSE}
data_for_JAGS <- list(
  n = dim(Y)[1],
  p = dim(Y)[2],
  q = q,
  Y = Y,
  X=X,
  F_x = ncol(X)
)

n.adapt = 1000
burnin = 1000
n.iter = burnin * 2
thin = 1
jm <- jags.model(
  file = textConnection(modelString),
  data = data_for_JAGS,
  n.chains = 3 ,
  n.adapt = n.adapt
)
update(jm, burnin)
jsamples <- coda.samples(
  model = jm,
  variable.names = c("beta","lambda","preciE"),
  n.iter = n.iter/10,
  thin = thin
)
```


We transform the MCMC coda file into a regular tibble to visualize posterior pdfs and variational ones.

```{r MCMCvisualization}
MCMC_results <- ggs(jsamples)
ggs_traceplot(MCMC_results)
ggs_density(MCMC_results,family=c("beta"))
ggs_caterpillar(MCMC_results,family = "beta") 

+
  scale_x_continuous(trans = "log10")
```

We can now visualize the estimated posterior distributions for $\beta$ parameters:
```{r}
MCMC_results <- ggs(jsamples)
MCMC_results %>% filter(str_detect(Parameter,"beta")) %>%
  group_by(Parameter) %>%  
  summarise(q50=mean(value), q25= quantile(value,probs=0.25),q75= quantile(value,probs=0.75),qinf= quantile(value,probs=0.25), qsup= quantile(value,probs=0.95) ) %>% 
  rename(names=Parameter) %>% cbind(color="Jags") ->Mjags

names=(levels(Mjags$names))
names<-names[str_detect(names,"beta")]

Mtruth<- tibble(names=names,q50=as.vector(t(beta_true))) %>% 
  mutate(q25=q50,q75=q50,qsup=q50,qinf=q50) %>% cbind(color="Truth")
  

meanligne= as.vector(t(result$params$Beta$M))
varligne=c()
for(f in 1:F_x){varligne=c(varligne,result$params$Beta$Cov[f,f,])}
MVI<- tibble(names=names,q50=meanligne) %>% 
  mutate(q25=q50+qnorm(0.25)*sqrt(varligne),
         q75=q50+qnorm(0.75)*sqrt(varligne),
         qsup=q50+qnorm(0.95)*sqrt(varligne),
         qinf=q50+qnorm(0.05)*sqrt(varligne)) %>% cbind(color="VI")
M=rbind(Mjags,Mtruth,MVI)
  g=ggplot(M,aes(x=names))+
  geom_boxplot(aes(color=color, ymin=qinf,lower=q25,middle=q50,
                  upper=q75,ymax=qsup),stat="identity",position = "dodge")+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
labs(y="posterior",x="beta")
print(g)



```

```{r}
n_samples <-n.iter/thin
result_VI<-list()
q_guess<-q
result_VI_coef<-result$params
# Faisons une structure similaire à Gibbs
result_VI$Lambda<-array(0,dim = c(p,q_guess,n_samples))
for(j in 1:p){result_VI$Lambda[j,1:q_guess,1:n_samples]=
  t(mvtnorm::rmvnorm(n_samples,result_VI_coef$Lambda$M[1:q_guess,j],
                     result_VI_coef$Lambda$Cov[1:q_guess,1:q_guess,j]))}
result_VI$Eta<-array(0,dim = c(n,q_guess,n_samples))
for(i in 1:n){result_VI$Eta[i,1:q_guess,1:n_samples]=
  t(mvtnorm::rmvnorm(n_samples,result_VI_coef$Eta$M[,i],
                     result_VI_coef$Eta$Cov[,,i]))}
result_VI$Sigma<-matrix(0,nr=p,nc=n_samples)
for(s in 1:n_samples){result_VI$Sigma[1:p,s]=1/rgamma(p,result_VI_coef$Sigma$A[1:p],
                                                      result_VI_coef$Sigma$B[1:p])}


Sigma_VI_df<- format_matrix(matrix_=result_VI$Sigma,
                            param_name="s",
                            suffix_ = "2")%>% 
  mutate(method="VI")

Lambda_hat_VI_df <- format_array(result_VI$Lambda,
                                 "Lambda") %>% mutate(method="VI")
```


