---
title: "About Variational Bayes for PPCA data"
author: "Eric PARENT & Pierre Gloaguen"
date: "the 24th of Mai 2022 , UMR MIA-Paris, AgroParisTech, INRA, Université Paris-Saclay"
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_document:
    code_folding: hide
    df_print: paged
    self_contained: yes
    number_sections: yes
    toc: yes
    theme: journal
    toc_float: yes
    toc_depth: 2
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE, cache = FALSE}

# knitr::opts_chunk$set(
# 	echo = TRUE,
# 	message = FALSE,
# 	warning = FALSE,
# 	cache = FALSE
# )
knitr::opts_chunk$set(
	eval = FALSE,
	message = FALSE,
	warning = FALSE,
	cache = FALSE,
	include = TRUE
)
```

# R codes and librairies {-}

All `R` codes can be unfolded, for instance, unfold the following chunk to see which libairies are used in this document:

```{r librairies}
rm(list=ls())
library(tidyverse) # For data manipulation and visualization
library(rjags)  # For MCMC sampling
library(ggmcmc) # To format jags output
library(extraDistr) # For the non standart t distribution
```


# A not-so-simple shrinkage PPCA model

Bhattacharya and Dunson consider the following linear normal model with a gamma process to shrink the variances of the columns of the latent matrix $\lambda$:
\begin{align*}
Y_i &\sim_{ind.} \mathcal{N}_p(\Lambda\eta_i, \Sigma)\;,\; 1\leq i \leq n \\
\Sigma &= diag(\sigma_j^{-2}),\; 1\leq j \leq p \\
\sigma_j^{-2}|(a_{\sigma},b_{\sigma}) &\sim \Gamma(a_{\sigma},b_{\sigma})\\
\lambda_{jh} &\sim \mathcal{N}_{q}(0,\,\phi_{jh}^{-1}\tau_{h}^{-1})
\; 1 \leq h \leq {q} \\
\phi_{jh} &\sim \Gamma(\frac{\nu}{2},\frac{\nu}{2})\\
\tau_{h} &= \prod_{l=1}^h \delta_l \\ 
\delta_1 &\sim \Gamma(a_1,1) \\
\delta_l &\sim \Gamma(a_2,1) \ \forall l \geq 2
\end{align*}

Recall that -- up to constant terms-- the logdensity of such a Normal multivariate distribution is:
$$\log[Y|\Lambda,\Sigma] = -\frac{n}{2}\left(log\vert C\vert +tr(C^{-1}\hat{S}_Y)\right)$$
with

$$ C= \Lambda\Lambda'+\Sigma\\
\hat{S}_Y=\frac{1}{n}\sum_{i=1}^n Y'_iY_i$$

In the case $q<<p$, because $\Sigma$ is diagonal, it is easier to compute the inverse of the variance-covariance matrix by $$C^{-1}=\Sigma^{-1}-\Sigma^{-1}\Lambda(I+ \Lambda' \Sigma^{-1}\Lambda)^{-1}\Lambda'\Sigma^{-1}$$


Some example data are generated in the following R code chunk.

```{r generation}
source("utils_PPCA_generating_data.R") # For true values
Y <- read.table("data_PPCA.txt", sep = ";") %>%
  as.matrix()
```


# Variationnal approach

We denote $Z$ the unknown with components $(\tau,\phi,\eta,\Lambda,\Sigma)$ and choose an independent family $\left(\prod_u q_{z_u}(Z_u) \right)$ to approximate $[Z|Y]$ such that :
\begin{align*}
\delta_h &\sim q_{\delta_h}(\cdot)= \mathcal{G}(A^{\delta}_h,B^{\delta}_h)\;,\; 1\leq h \leq q \\
\phi_{jh} &\sim q_{\phi_{jh}}(\cdot)= \mathcal{G}(A^{\phi}_{jh},B^{\phi}_{jh})\;,\; 1\leq j \leq p \;,\; 1\leq h \leq q\\
\sigma^{-2}_{j} &\sim q_{\sigma^{-2}_{j}}(\cdot)= \mathcal{G}(A^{\sigma}_j,B^{\sigma}_j)\;,\; 1\leq j \leq p \\
\eta_i &\sim q_{\eta_i}(\cdot)= \mathcal{N}(M_{\eta_i},(V^{\eta_i}))\;,\; 1\leq i \leq n \\
\lambda_j &\sim q_{\lambda_j }(\cdot)= \mathcal{N}(M_{\lambda_j},(V^{\lambda_j}))\;,\; 1\leq j \leq p 
\end{align*}
We rely on the minimisation of the Kullback Liebler divergence
$KL= \mathbb{E}_q\left[\log\frac{q(Z}{[Z|Y]}\right]$.
Equivalently we want to maximize the ELBO

$$ELBO= \mathbb{E}_q\left[\log[Y,\tau,\phi,\eta,\Lambda,\sigma^{-2}]\right]-\mathbb{E}_q\left[\log q(\tau,\phi,\eta,\Lambda,\sigma^{-2})\right]$$

Due to conditionnal independence the criterion can decomposed as 

\begin{align*}
ELBO=&\mathbb{E}_q\left[\log\left([Y|\eta,\Lambda,\Sigma]\right)\right]\\
& + \mathbb{E}_q\left[\log\left([\Sigma]\right)\right]\\
& + \mathbb{E}_q\left[\log\left([\eta]\right) \right]\\
& + \mathbb{E}_q\left[\log\left([\Lambda|\delta,\phi]\right)\right]\\
& + \mathbb{E}_q\left[\log\left([\phi]\right)\right]\\
& + \mathbb{E}_q\left[\log\left([\delta]\right)\right]\\
& -\mathbb{E}_q\left[\log q(\Lambda,\Sigma,\eta,\phi,\delta)\right]
\end{align*}

Up to constant terms:
$$\log[Y|\Lambda,\Sigma,\eta] = -\frac{n}{2}\left(log\vert \Sigma^{-1}\vert +tr(\Sigma^{-1}\hat{S}_{Y,\eta,\Lambda})\right)$$
with

$$ \hat{S}_{Y,\eta,\Lambda}=\frac{1}{n}\sum_{i=1}^n (Y_i-\Lambda\eta_i')(Y'_i-\eta_i\Lambda')\\
= \frac{1}{n} (Y'-\eta\Lambda')(Y-\lambda\eta')$$


and 
$$\log[\eta] = -\frac{n}{2}\left(log\vert I_p\vert +tr(I_p\hat{S}_{\eta})\right)$$
with

$$ \hat{S}_{\eta}=\frac{1}{n}\sum_{i=1}^n \eta_i\eta_i'\\
 = \frac{1}{n} \eta'\eta$$
 

Since $\Sigma$ is diagonal:

$$\log[\Sigma^{-1}] = \sum_{j=1}^p\left((a_\sigma -1)log(\sigma_j^{-2}) -b_\sigma\sigma_j^{-2}\right)$$

Due to normal assumptions for $\Lambda$

$$\log\left([\Lambda_{jh}|\tau_{h},\phi_{jh}]\right)=0.5log(\phi_{jh})-0.5\Lambda_{jh}^2\phi_{jh}\prod_{k=1}^h \delta_k+0.5\sum_{k=1}^h log(\delta_k) $$
with gamma shrinkage process for the precision

$$\log[\phi_{jh}] = (\frac{1}{ 2} \gamma -1)log(\phi_{jh}) -\frac{1}{ 2}\phi_{jh}$$
$$\log[\delta_{1}] = (a_1 -1)log(\delta_{1}) -b_{1} \delta_{1}\\
\log[\delta_{2}] = (a_2 -1)log(\delta_{2}) -b_{2} \delta_{2}\\
\cdots \\
\log[\delta_{h}] = (a_2 -1)log(\delta_{h}) -b_{2} \delta_{h} \; , \; h\ge 2$$

Wrapping up all the previous log density terms


\begin{align*}
ELBO=& -\mathbb{E}_q\left[\log q(\Lambda,\Sigma,\eta,\phi,\delta)\right]\\
& +\mathbb{E}_q\left[\frac{n}{2}\left(\sum_{j=1}^p \log\sigma_j^{-2} \right)-0.5\sum_{i=1}^n tr\left(\Sigma^{-1} (Y_i-\Lambda\eta_i')(Y'_i-\eta_i\Lambda')\right)\right]\\
& + \mathbb{E}_q\left[\sum_{j=1}^p\left((a_\sigma -1)log(\sigma_j^{-2}) -b_\sigma\sigma_j^{-2}\right)\right]\\
& + \mathbb{E}_q\left[\frac{1}{2}tr(\eta'\eta) \right]\\
& + \sum_{j=1,h=1}^{p,q}\mathbb{E}_q\left[0.5log(\phi_{jh})-0.5\Lambda_{jh}^2\phi_{jh}\prod_{k=1}^h \delta_k+0.5\sum_{k=1}^h log(\delta_k) \right]\\
& + \mathbb{E}_q\left[(\frac{1}{ 2} \gamma -1)log(\phi_{jh}) -\frac{\gamma}{ 2}\phi_{jh}\right]\\
& + \mathbb{E}_q\left[(a_1 -1)log(\delta_{1}) - \delta_{1}\right]\\
& + \mathbb{E}_q\left[(a_2 -1)log(\delta_{2}) - \delta_{2}\right]\\
& + \cdots\\
& + \mathbb{E}_q\left[(a_2 -1)log(\delta_{q}) - \delta_{q}\right]
\end{align*}


Note that in all cases, expectations are  analytically available.

A convenient way to maximize this function is to proceed iteratively via a coordinate ascent algorithm. 

Hence

 * $A^{\sigma}_j= a_\sigma+\frac{n}{2}$; 
 * $B^{\sigma}_j= b_\sigma+\frac{1}{2}\sum_{i=1}^n \mathbb{E}_q ((Y_{ij}-\Lambda'_j\eta_i)^2)$; when developping the expression we get: $\mathbb{E}_q ((Y_{ij}-\Lambda'_j\eta_i)^2)= \mathbb{E}_q (Y_{ij}^2)-2\mathbb{E}_q (Y_{ij})\mathbb{E}_q (\Lambda'_j\eta_i)+\sum_{h,h'}^{q,q} \mathbb{E}_q (\Lambda'_{jh}\Lambda'_{jh'}) \mathbb{E}_q (\eta_{i,h}\eta_{i,h'})$. The last term is the sum of the elements of the matrix resulting from a term-by-term multiplication of covariance matrices. Hence,  $\sum_{h,h'}^{q,q} \mathbb{E}_q (\Lambda'_{jh}\Lambda'_{jh'}) \mathbb{E}_q (\eta_{i,h}\eta_{i,h'})= \sum_{h,h'}^{q,q} \left(\mathbb{E}_q (\Lambda_{j})\mathbb{E}_q (\Lambda'_{j})+ \mathbb{V}_q (\Lambda_{j})\right) .\times. \left(\mathbb{E}_q (\eta_{i})\mathbb{E}_q (\eta'_{i})+ \mathbb{V}_q (\eta_{i})\right)$
 
 * $A^{\phi}_{jh}= \frac{\gamma}{2}+\frac{1}{2}$; 
 * $B^{\phi}_{jh}=\frac{\gamma}{2}+\frac{1}{2}\mathbb{E}_q (\Lambda_{jh} ^2 ) \prod_{k=1}^h \mathbb{E}_q (\delta_k)$; 
 
 * $A^{\delta}_1= a_1+\frac{p}{2}$; 
 * $B^{\delta}_{1}=1+\frac{1}{2}\sum_{l=1}^q \sum_{j=1}^p \mathbb{E}_q (\Lambda_{jl} ^2 )\mathbb{E}_q (\phi_{jl} ) \prod_{t=2}^l \mathbb{E}_q (\delta_t)$;  
 
 for $l \ge 2$: 
 
 * $A^{\delta}_h= a_2+\frac{p(q-h+1)}{2}$; 
 * $B^{\delta}_{h}=1+\frac{1}{2}\sum_{l=h}^q \sum_{j=1}^p \mathbb{E}_q (\Lambda_{jl} ^2 )\mathbb{E}_q (\phi_{jl} ) \prod_{t=1, t\ne h}^l \mathbb{E}_q (\delta_t)$;  

 
 The variational normal distribution for $\eta_i$ is obtained as follows:
 
 * First, let's compute the precision first component, $\tilde{P}^{\eta_i}=\sum_{j=1}^p \frac{A^{\sigma_j}}{B^{\sigma_j}}V^{\lambda_j}$
 * Then the precision, $(V^{\eta_i})^{-1}=\tilde{P}^{\eta_i}+I_{qq}$
 
 * And subsequently the variational mean
 $M^{\eta_i}=V^{\eta_i}\mathbb{E}_q (\Lambda')\mathbb{E}_q ( \Sigma^{-1})Y_i$
 
 The variational normal distributions for $\lambda_j$ is obtained as follows:
 
 * ${P}^{\lambda_j}=\mathbb{E}_q({\sigma_j^{-2}}\eta'\eta)+\mathbb{E}_q({D^{-1}}_\eta'\eta_j)$
 
 * The precision first component, $\tilde{P}^{\lambda_j}=\left(\frac{A^{\sigma_j}}{B^{\sigma_j}}\right)\sum_{i=1}^n (V^{\eta_i}+{M^{\eta_i}}{M^{\eta_i}}')$ . Note that the matrix $\sum_{i=1}^n (V^{\eta_i}+{M^{\eta_i}}{M^{\eta_i}}')$ is to be computed once for all $1\le j \le p)$
 
 * Then the precision, $(V^{\lambda_j})^{-1}=\tilde{P}^{\lambda_j}+diag(\frac{A^{\phi_{jh}}}{B^{\phi_{jh}}} \prod_{l=1}^h\frac{A^{\delta_{l}}}{B^{\delta_{l}}} )$
 
 * And subsequently the variational mean
 $M^{\lambda_j}=V^{\lambda_j}\left(\frac{A^{\sigma_j}}{B^{\sigma_j}}\right) {M^{\eta}}'Y_j$
 
 These update equations are implemented within functions.
 
We can check the evolution of the ELBO across iterations. To compute the ELBO, one needs to recall that

 * the entropy of a Normal ($\mu,\Sigma$) distribution is $\frac{log(\vert \Sigma \vert)}{2}+ \frac{dim(\Sigma)}{2} (1+log(2\pi))$;
 
 * the entropy of a Gamma ($\alpha,\beta$) distribution is $\alpha-log(\beta)+log\Gamma(\alpha)+(1-\alpha) \psi(\alpha)$ with $\psi(.)$ being the so-call digamma function.

Hence: 
\begin{align*}
ELBO=& -\mathbb{E}_q\left[\log q(\Lambda,\Sigma,\eta,\phi,\delta)\right]\\
& +\frac{n}{2}\sum_{j=1}^p\left( \psi(A_j^{\sigma_j})-log(B_j^{\sigma_j}) \right)\\
& +\mathbb{E}_q\left[-0.5\sum_{i=1}^n tr\left(\Sigma^{-1} (Y_i-\Lambda\eta_i')(Y'_i-\eta_i\Lambda')\right)\right]\\
& + \mathbb{E}_q\left[\sum_{j=1}^p\left((a_\sigma -1)log(\sigma_j^{-2}) -b_\sigma\sigma_j^{-2}\right)\right]\\
& + \mathbb{E}_q\left[\frac{1}{2}tr(\eta'\eta) \right]\\
& + \sum_{j=1,h=1}^{p,q}\mathbb{E}_q\left[-0.5\Lambda_{jh}^2\frac{A_{jh}^{\phi}} {B_{jh}^{\phi}}\prod_{k=1}^h \frac{A_k^{\delta}} {B_k^{\delta}}+0.5\sum_{k=1}^h \left( \psi(A_k^{\delta})-log(B_k^{\delta}) \right) \right]\\
& + (\frac{1}{ 2} \gamma -\frac{1}{ 2})\left( \psi(A_{jh}^{\phi})-log(B_{jh}^{\phi}) \right) -\frac{\gamma}{ 2}\frac{A_{jh}^{\phi}} {B_{jh}^{\phi}}\\
& +  (a_1 -1)\left( \psi(A_1^{\delta})-log(B_1^{\delta}) \right) -\frac{A_1^{\delta}} {B_1^{\delta}}\\
& +  (a_2 -1)\left( \psi(A_2^{\delta})-log(B_2^{\delta}) \right) -\frac{A_2^{\delta}} {B_2^{\delta_q}}\\
& + \cdots\\
& + (a_2 -1)\left( \psi(A_q^{\delta})-log(B_q^{\delta}) \right) -\frac{A_q^{\delta}} {B_q^{\delta}}
\end{align*}
 

Before exploring results, we will perform the inference on the same model with MCMC (in `Jags`).

# Comparaison with MCMC inference

Relying on Jags, we can get MCMC samples from the posterior distribution of the unknows.

```{r modelString}
modelString = "
model{
#a1 <- 2
#a2 <- 3
a[1] ~ dgamma(2,1)T(2,)
a[2] ~ dgamma(2,1)T(3,)
deminu <- 3/2
a_sigma <- 0.1
b_sigma <- 0.1

#shrinkage prior
for( h in 2:q){delta[h] ~ dgamma(a[2],1)}
              delta[1] ~ dgamma(a[1],1)
for( h in 1:q){ tau[h] <- prod(delta[1:h])} 
for( j in 1:p){
               for( h in 1:q){
                              phi[j,h] ~ dgamma(deminu,deminu)
                              precilambda[j,h] <- phi[j,h]*tau[h]
                              lambda[j,h]~ dnorm(0,precilambda[j,h])
                              }
               }
# N(0,1) prior for eta
for(i in 1:n){
              for (h in 1:q){
                      eta[i,h] ~ dnorm(0,1)
                            }
}
muY <- eta %*% t(lambda)
# finally Y
for(j in 1:p){ preciE[j] ~ dgamma(a_sigma,b_sigma)
              for (i in 1:n){
                      Y[i,j] ~ dnorm(muY[i,j],preciE[j])
                            }
}

}"
```


```{r jags_inference, eval = FALSE}
data_for_JAGS <- list(
  n = dim(Y)[1],
  p = dim(Y)[2],
  q = 6,
  Y = Y
)
n.adapt = 1000
burnin = 1000
n.iter = burnin * 2
thin = 1
jm <- jags.model(
  file = textConnection(modelString),
  data = data_for_JAGS,
  n.chains = 3 ,
  n.adapt = n.adapt
)
update(jm, burnin)
jsamples <- coda.samples(
  model = jm,
  variable.names = c("tau","a"),
  n.iter = n.iter,
  thin = thin
)
```


We transform the MCMC coda file into a regular tibble to visualize posterior pdfs and variational ones.

```{r MCMCvisualization}
MCMC_results <- ggs(jsamples)
ggs_traceplot(MCMC_results)
ggs_density(MCMC_results %>%
              filter(Parameter %in% c("a[1]","a[2]")))
ggs_density(MCMC_results,family=c("tau"))
ggs_caterpillar(MCMC_results,family = "tau") +
  scale_x_continuous(trans = "log10")
```

We can now visualize the estimated posterior distributions for all parameters:


